---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(width = 120)
```

# dtlg <a href="https://AscentSoftware.github.io/dtlg"><img src="man/figures/logo.svg" align="right" height="139" alt="dtlg website" /></a>

<!-- badges: start -->
[![CRAN version](https://www.r-pkg.org/badges/version/dtlg)](https://cran.r-project.org/package=dtlg)
[![R-CMD-check](https://github.com/AscentSoftware/dtlg/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/AscentSoftware/dtlg/actions/workflows/R-CMD-check.yaml)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://cran.r-project.org/web/licenses/MIT)
[![CRAN downloads last month](http://cranlogs.r-pkg.org/badges/last-month/dtlg)](https://cran.r-project.org/package=dtlg)
<!-- badges: end -->

`{dtlg}` builds the summary tables that populate Section 14 of a Clinical Study
Report --- demographics, adverse event incidence, and laboratory summaries ---
from ADaM-like analysis datasets. The compute backend is data.table, fast enough
for interactive Shiny dashboards on populations of $10^5$ to $10^6$
subjects. Constructors return plain data.tables, ready to render with
kableExtra, gt, DT, or any other formatter.

## Installation

``` r
install.packages("dtlg")
```

## Example

Creating an AET01 table:

```{r}
library(dtlg)

head(adsl[, c("USUBJID", "ARM")])

head(aesi[, c("USUBJID", "ARM", "FATAL")])

AET01 <-
  AET01_table(
    adsl = adsl,
    adae = aesi,
    patient_var = "USUBJID",
    treat_var = "ARM",
    aesi_vars = c("FATAL", "SER", "SERWD", "SERDSM", "RELSER",
                  "WD", "DSM", "REL", "RELWD", "RELDSM", "SEV")
  )

AET01[]
```

