---
title: "Getting started with dtlg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dtlg}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
```

```{r setup, message = FALSE}
library(dtlg)
```

## What dtlg is for

`{dtlg}` builds the summary tables that populate Section 14 of a Clinical
Study Report (CSR): demographics, adverse event incidence, and laboratory
summaries. The package covers the tables in the TLG triad; listings and
graphs are out of scope. Inputs are ADaM-like analysis datasets (ADSL,
ADAE, ADLB); outputs are tidy `data.table`s ready to render with any
downstream formatter (`kableExtra`, `gt`, `flextable`, `rtables`, or a Shiny
widget).

The package began as a response to a specific frustration: Shiny dashboards
for clinical review needed aggregations fast enough to recompute on user
input. The reference implementations in `{tern}` and `{rtables}` are
authoritative but were not designed for the round-trip latency of an
interactive session. Two design choices follow from that origin:

- The compute backend is `data.table`. Aggregation over populations of
  $10^5$ to $10^6$ subjects remains interactive. See
  `vignette("benchmarks")` for comparisons against `{tern}`.

- The output is always a plain `data.table` with a `stats` column and one
  column per treatment level. There is no bespoke table object, no print
  method to learn, and no rendering coupled to the computation. Indent
  strings default to HTML non-breaking spaces (`nbsp()`) so that the same
  output renders correctly in a Shiny `renderTable()` or `renderDT()`
  call without re-formatting.

This article walks through the user-facing API by reproducing three
representative tables.

## Example data

The package re-exports four ADaM-shaped datasets from `random.cdisc.data` for
illustration and testing:

| Object | Description                                                  |
|--------|--------------------------------------------------------------|
| `adsl` | Subject-level analysis dataset (one row per subject).        |
| `adae` | Adverse events analysis dataset.                             |
| `aesi` | `adae` filtered to analysis records, with derived AESI flags.|
| `adlb` | Laboratory measurements in BDS format.                       |

```{r}
adsl[1:3, c("USUBJID", "ARM", "AGE", "SEX", "RACE")]
```

## A first table: demographics

`summary_table()` is the high-level entry point. It dispatches on the type of
each `target` variable: numeric columns are summarised with `calc_desc()`
(n, mean (SD), median, min/max, missing); categorical columns are summarised
with `calc_counts()` (n (%) per level, denominator from `.total_dt`).

The `indent` argument defaults to `nbsp(n = 4L)` --- four HTML non-breaking
spaces --- which renders correctly in Shiny and any HTML formatter passed
`escape = FALSE`. For console inspection, pass `indent = "  "` so the output
is legible without HTML rendering:

```{r}
dmg_vars <- c("AGE", "SEX", "RACE", "BMRKR1")
dmg_lbls <- c("Age (yr)", "Sex", "Race", "Biomarker 1")

dm_table <- summary_table(
  adsl,
  target      = dmg_vars,
  target_name = dmg_lbls,
  treat       = "ARM",
  indent      = "  "
)

dm_table
```

The first column carries the variable heading and the indented statistic
labels. The remaining columns are the levels of `treat`. Reorder columns
explicitly with `treat_order`:

```{r}
summary_table(
  adsl,
  target      = "AGE",
  target_name = "Age (yr)",
  treat       = "ARM",
  treat_order = c("B: Placebo", "A: Drug X", "C: Combination"),
  indent      = "  "
)
```

## Building blocks for event tables

Safety tables are constructed from small, composable primitives. Each
returns a one-element list wrapping a `data.table`; `merge_table_lists()`
unwraps and row-binds them. The list wrapping exists so that mixed-shape
intermediates can be assembled without intermediate casting.

| Helper                | Counts                                                |
|-----------------------|-------------------------------------------------------|
| `event_count()`       | Patients matching a predicate (e.g. `DTHFL == 'Y'`).  |
| `total_events()`      | Records, not patients (e.g. total AEs).               |
| `multi_event_true()`  | Patients flagged across many binary indicators.       |
| `event_count_by()`    | Patients and events nested by a grouping variable.    |

A minimal example. Deaths per arm:

```{r}
event_count(
  adsl,
  patient  = "USUBJID",
  treat    = "ARM",
  label    = "Total number of deaths",
  .filters = "DTHFL == 'Y'"
)[[1]]
```

`.filters` is a character vector of unevaluated R expressions; they are
parsed and applied in the frame of `dt`. Passing multiple expressions
combines them with logical AND.

## Composing AET01 from primitives

The AET01 table is the canonical safety overview. The high-level
`AET01_table()` wraps the composition shown below; understanding the
primitives makes the result auditable.

```{r}
aesi_vars <- c(
  "FATAL", "SER", "SERWD", "SERDSM", "RELSER",
  "WD", "DSM", "REL", "RELWD", "RELDSM", "SEV"
)

deaths <- event_count(
  adsl,
  patient  = "USUBJID",
  treat    = "ARM",
  label    = "Total number of deaths",
  .filters = "DTHFL == 'Y'"
)

withdrawals <- event_count(
  adsl,
  patient  = "USUBJID",
  treat    = "ARM",
  label    = "Total number of patients withdrawn due to an AE",
  .filters = "DCSREAS == 'ADVERSE EVENT'"
)

patients_any_ae <- event_count(
  aesi,
  patient   = "USUBJID",
  treat     = "ARM",
  label     = "Total number of patients with at least one AE",
  .total_dt = adsl
)

total_ae <- total_events(
  aesi,
  treat = "ARM",
  label = "Total number of AEs"
)

aesi_block <- multi_event_true(
  aesi,
  event_vars = aesi_vars,
  patient    = "USUBJID",
  treat      = "ARM",
  heading    = "Total number of patients with at least one",
  .total_dt  = adsl,
  indent     = "  "
)

aet01 <- merge_table_lists(list(
  patients_any_ae,
  total_ae,
  deaths,
  withdrawals,
  aesi_block
))

aet01
```

Two details worth noting. First, `.total_dt = adsl` supplies the denominator
for percentages; without it the denominator is `dt` itself, which silently
yields the wrong result whenever the analysis dataset is filtered.
Second, the AESI block uses the `label` attribute of each flag variable for
row labels when `label` is not supplied explicitly; `with_label()` sets
that attribute.

The one-liner equivalent is `AET01_table()`:

```{r}
aet01_oneliner <- AET01_table(
  adsl        = adsl,
  adae        = aesi,
  patient_var = "USUBJID",
  treat_var   = "ARM",
  aesi_vars   = aesi_vars
)

identical(aet01, aet01_oneliner)
```

## AE incidence by SOC and PT

`AET02_table()` produces the System Organ Class / Preferred Term breakdown,
combining `event_count()`, `total_events()`, and `event_count_by()`:

```{r}
aet02 <- AET02_table(
  adsl    = adsl,
  adae    = aesi,
  patient = "USUBJID",
  treat   = "ARM",
  target  = "AEDECOD",
  rows_by = "AEBODSYS",
  indent  = "  "
)

head(aet02, 12)
```

## Longitudinal laboratory summaries

For BDS-shaped data such as `adlb`, `summary_table_by()` groups rows by one
or more nesting variables, and `summary_table_by_targets()` summarises two
target columns side by side --- the typical layout for value and
change-from-baseline:

```{r}
adlb_post <- adlb[adlb$AVISIT != "SCREENING", ]

lb_table <- summary_table_by_targets(
  dt      = adlb_post,
  target  = c("AVAL", "CHG"),
  treat   = "ARM",
  rows_by = c("PARAM", "AVISIT"),
  indent  = "  "
)

head(lb_table, 8)
```

The output column names are suffixed with `.AVAL` and `.CHG` so the two
blocks remain distinguishable after the column-bind.

## Output is a plain data.table

Every constructor returns a `data.table`. Rendering is the caller's choice.
Because the default indent and the formatted percentage strings contain
HTML entities (`&nbsp;`), pass `escape = FALSE` (or the equivalent) to the
renderer so the indentation is honoured:

```{r, eval = requireNamespace("kableExtra", quietly = TRUE)}
dm_html <- summary_table(
  adsl,
  target      = dmg_vars,
  target_name = dmg_lbls,
  treat       = "ARM"
)

dm_html |>
  kableExtra::kable(format = "html", escape = FALSE) |>
  kableExtra::kable_styling(full_width = FALSE)
```

In a Shiny app the same object goes straight into `renderTable(sanitize.text.function = identity)` or `DT::renderDT(escape = FALSE)`; no
intermediate conversion is required. This is the use case the package was
shaped around: a treatment-arm selector or a population filter recomputes
the table on every input change, and the round-trip stays interactive on
ADaM datasets at trial-realistic sizes.

For console inspection, `print_dtlg()` wraps `data.table`'s print method with
left-justified columns and no truncation:

```{r}
print_dtlg(head(aet01, 5))
```

## Copy semantics

`{dtlg}` exposes the `data.table` reference-vs-value trade-off explicitly
rather than silently choosing for the user. Internally, every constructor
calls `maybe_copy_dt()`, which consults the global option
`dtlg_dt_copy_semantics`:

- `"reference"` (default): a `data.frame` input is converted in place via
  `setDT()`. Subsequent operations alias the caller's object. This is fastest
  and matches `data.table` idioms.
- `"value"`: inputs are deep-copied. Constructors never mutate the caller's
  data, at the cost of one allocation per call.

Inspect and switch:

```{r}
dt_copy_semantics()
old <- set_dt_copy_semantics("value")
dt_copy_semantics()
set_dt_copy_semantics(old)
```

If you write `{dtlg}` calls inside a function that should preserve
`data.frame`-like inputs, set `"value"` semantics for the scope. For
production pipelines on large analysis datasets where the inputs are
already `data.table`s and not reused downstream, the default is correct
and avoids redundant allocation.

## Comparison with tern

`{tern}` and `{rtables}` produce the canonical reference output for these
tables. `{dtlg}` ships thin wrappers --- `tern_summary_table()`,
`tern_AET01_table()`, `tern_AET02_table()` --- that build the equivalent
`TableTree` through `{tern}` using the same call signature as the
`{dtlg}` constructor. `as_dtlg_table()` flattens a `TableTree` to the
`data.table` shape used in this article. Together they enable structured
side-by-side checks during validation:

```{r, eval = requireNamespace("tern", quietly = TRUE) && requireNamespace("rtables", quietly = TRUE)}
tt <- tern_summary_table(
  adsl,
  target      = c("AGE", "SEX"),
  target_name = c("Age (yr)", "Sex"),
  treat       = "ARM"
)

as_dtlg_table(tt)
```

## Where to go next

- `vignette("benchmarks")` --- timing comparisons against `{tern}` on
  realistic population sizes.
- The function reference, grouped by responsibility (Statistics, Events,
  Summary Tables, Clinical Tables, tern interop, Copy semantics).
- `?aesi` --- documentation of the derived AESI flags used throughout the
  safety examples.
